Memes: an R interface to the MEME Suite

### 2020-10-19 manuscript.Rmd
Merge changes w/ dan edits

# Bioconductor Abstract (300 words)

@@

# Abstract Draft (OLD: delete??)
Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools that implement these features have
historically existed in the Bioconductor landscape, they often rely on
proprietary data types with poor interoperability with other tools. Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching. Here we present `memes`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

`memes` uses base R and Bioconductor data types for better integration with
common analysis tools like the `tidyverse` and other Bioconductor packages.
Unlike the commandline implementation, `memes` outputs also function as inputs
to other MEME Suite tools, allowing simple construction of motif analysis
pipelines.

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression in the tissue
of interest. `memes` also provides simple utilities for data visualization where
appropriate.

@@

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis such as *de-novo* motif discovery, motif enrichment
testing, motif comparison, and motif matching, among others \cite{meme-suite}.
MEME Suite tools take as input fasta files of sequences which will be searched
for motifs, and if necessary, .meme format motif files containing motifs to use
in the search.

Data from MEME Suite tools can be underutilized because HTML interface?
 - HTML is great human-readable reporting interface
 - `memes` allows customized reporting.

Bioconductor is deprecating PWMEnrich (~AME) and MotIV (~TomTom), `memes` can fill the gap and add additional value.

Biopolymers, such as DNA and protein molecules, perform varying functions based
on the composition of their primary sequence. Short, repeated sequences, or
"motifs" represent functional units within biopolymers which can act as
interaction surfaces, create structure, or have enzymatic activity.
Identification of shared motifs across multiple sequences can provide evidence
for shared function, thus the ability to identify, classify, and compare motifs
represents a key step in the analysis of biological sequences. Although tools
that implement these features have historically existed in the Bioconductor
landscape, they often rely on proprietary data types with poor interoperability
with other tools. Furthermore, several of these packages will be deprecated in
the next Bioconductor release, leaving a gap in functionality with no clear
alternatives.

>>>>

# Abstract

@@

@@

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools that implement these features have
historically existed in the Bioconductor landscape, they often rely on
proprietary data types with poor interoperability with other tools. Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.
